The Essence of Summarization\\ in Pushdown Flow Analyses\\ \textbf{Supplemental Material}

### 2014-02-24 content.tex
Small changes. About to remove gratuitous parentheses.

%%
This paper gives a common, simple framework to derive CFA2, PDCFA and its garbage-collecting extension in a manner similar to AAM.
%
To exercise the framework further, we give a novel analysis for delimited, composable control. \\

@@

Constructing models and analyzing models are different processes.
%
The goal of this work is to simplify the program $\to$ model process.
%
First-order languages have the luxury of a fully-precise control flow graph that can be extracted from a program AST in linear time.
%
In a sense, the first-order world gets their models ``for free.''
%
Higher-order recursion schemes (HORS) are a powerful model for simply typed functional programs with good model-checking properties~\citep{dvanhorn:Ong2006ModelChecking}, but the program $\to$ HORS process is narrowly applicable.
%
It is still a matter of mystery for HORS to support programming language features like mutable state and untyped computation.
%%
As a middle ground, we show a simple way to transform one's programming language semantics into a what can be viewed as a pushdown model constructor.
%%

%%
When striving for precision, the interconnectedness of data-flow and control-flow analysis in higher-order languages extend upwards to the interconnectedness of model construction and model-checking.
%
When we tune the abstractions, the result of translating semantics can range from control-flow analysis~\citep{dvanhorn:VanHorn2010Abstracting} to a symbolic execution engine~\citep{dvanhorn:TobinHochstadt2012Higherorder}.
%
These are all forward-executing techniques, so the analogy does not fully extend to all analytical technologies.
%
We find the power and simplicity in the constructions enticing nonetheless.
%
The resulting artifacts can be used as models in the first-order tools~\citep{ianjohnson:bandera}

@@

CFA2 is an online version of a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with PDCFA's crucial ``$\epsilon$-closure graph''.
%
The online quality of the algorithm is what makes it suitable for higher-order languages.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
We can adopt the technique for higher precision in the common case but still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call/cc} operator in~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model like 0-CFA would provide.
%%

@@

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mexpr \in \Expr = \svar{\mvar} \alt \sapp{\mexpr}{\mexpr} \alt \slam{\mvar}{\mexpr}
\end{align*}
\begin{minipage}[b]{.55\linewidth}
  \begin{align*}
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \vclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \alt \kcons{\mkframe}{\mkont} \\
  \mkframe \in \Frame &::= \kar{\mexpr,\menv} \alt \kfn{\mval}
  \end{align*}\hrule height 0pt\end{minipage}
\begin{minipage}[b]{.40\linewidth}
  \begin{align*}
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
  \end{align*}\hrule height 0pt\end{minipage}
}
%%

%%
The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\alloc(\tpl{(\mexpr,\menv),\mstore,\mkont}) \notin \dom(\mstore)$), but any strategy is sound~\citep{dvanhorn:Might2009Posteriori}.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of continuation frames is finite.
%
A finite allocation strategy makes the continuation the only unbounded data structure in the machine, which follows a stack discipline with a finite ``stack alphabet''---thus the pushdown system interpretation is apt.
%
The $\alloc$ meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store.
%
We borrow this technique on top of our own.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kcons{\kfn{\mval}}{\mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine with allocation tuning}
  \label{fig:base-semantics}
\end{figure}

>>>>

%%
This paper gives a common, simple methodology to derive CFA2, PDCFA and its garbage-collecting extension in a manner similar to AAM.
%
Our presentation is meant to be a tutorial to this related literature, with the additional promise of a novel analysis.
%
We don't presume a language of formal semantics such that our technique applies to all formalisms in such a language.
%
Such an approach is unlikely to capture all ``real language's'' formalizations.
%
An additional abstraction also detracts from the operational intuitions and simplicity of the methodology.

@@

Constructing models and analyzing models are different processes.
%
The goal of this work is to simplify the \textit{program} $\to$ \textit{model} process.
%
First-order languages have the luxury of a fully-precise control flow graph that can be extracted from a program AST in linear time.
%
In a sense, the first-order world gets their models ``for free.''
%
Higher-order recursion schemes (HORS) are a powerful model for simply typed functional programs with good model-checking properties~\citep{dvanhorn:Ong2006ModelChecking}, but the \textit{program} $\to$ \textit{HORS} process is narrowly applicable.
%
It is still a matter of mystery for HORS to support programming language features like mutable state and untyped computation.
%%
As a middle ground, we show a simple way to transform one's programming language semantics into a what can be viewed as a pushdown model constructor.
%%

%%
When striving for precision, the interconnectedness of data-flow and control-flow analysis in higher-order languages extends upwards to the interconnectedness of model construction and model-checking.
%
When we tune the abstractions, the result of translating semantics can range from control-flow analysis~\citep{dvanhorn:VanHorn2010Abstracting} to a symbolic execution engine~\citep{dvanhorn:TobinHochstadt2012Higherorder}.
%
These are all forward-executing techniques, so the analogy does not fully extend to all analytical technologies.
%
We find the power and simplicity in the constructions enticing nonetheless.
%
The resulting artifacts can be used as models in the first-order tools, as demonstrated by the Java model-checker Bandera~\citep{ianjohnson:bandera}.

@@

CFA2 is an online version of a technique called ``summarization'' from Sharir and Pnueli \citep[Chapter 7]{local:muchnick:jones:flow-analysis:1981}, which is synonymous with PDCFA's crucial ``$\epsilon$-closure graph''.
%
The online quality of the algorithm is what makes it suitable for higher-order languages.
%
Summarization algorithms need not be restricted to languages with well-bracketed calls and returns.
%
The technique is adaptable for higher precision in the common case but to still handle difficult cases such as first-class control.
%
This was shown for the \rackett{call/cc} operator in~\citet{ianjohnson:Vardoulakis2011Pushdown}.
%
This work illuminated the fact that we can harness the enhanced technology of pushdown analyses in non-pushdown models of computation.
%
Doing this sacrifices call/return matching in the general case, but in practice the precision is much better than the alternative regular model like 0-CFA would provide.
%%

@@

{
\setlength{\abovedisplayskip}{0pt}
\setlength{\belowdisplayskip}{4pt}
\setlength{\abovedisplayshortskip}{0pt}
\setlength{\belowdisplayshortskip}{8pt}
\begin{align*}
  \mexpr \in \Expr &= \svar{\mvar} \alt \sapp{\mexpr}{\mexpr} \alt \slam{\mvar}{\mexpr} \\
  \mstate \in \State &= (\Expr \times \Env) \times \Store \times \Kont \\
  \mval \in \Value &::= \vclo{\slam{\mvar}{\mexpr}}{\menv} \\
  \mkont \in \Kont &::= \kmt \alt \kcons{\mkframe}{\mkont} \\
  \mkframe \in \Frame &::= \kar{\mexpr,\menv} \alt \kfn{\mval} \\
  \menv \in \Env &= \Var \to \Addr \\
  \mstore \in \Store &= \Addr \to \wp(\Value) \\
  \mvar \in \Var &\text{ an infinite set} \\
  \maddr \in \Addr &\text{ an infinite set}
\end{align*}
}
%%

%%
%
The $\kmt$ continuation is the empty continuation.
%
The continuation frames represent an \textbf{ar}gument left to evaluate, and \textbf{f}u\textbf{n}ction left to call.
%
%
The $\Addr$ space is what controls the precision of the model.
%
For a concrete semantics, we require the allocation meta-function $\alloc :
\State \to \Addr$ to return fresh addresses ($\alloc(\tpl{(\mexpr,\menv),\mstore,\mkont}) \notin \dom(\mstore)$), but any strategy is sound~\citep{dvanhorn:Might2009Posteriori}.
%
If $\alloc$ only uses addresses from a finite subset of $\Addr$, then the state space without continuations is finite, and the space of continuation frames is finite.
%
A finite allocation strategy makes the continuation the only unbounded data structure in the machine.
%
The semantics' treatment of continuations follows a stack discipline, and since it has a finite ``stack alphabet'' (space of frames), a pushdown system interpretation is apt.
%
The $\alloc$ meta-function along with the commitment to having no recursive data-structures is the key to the AAM technique.
%
All recursion can be expressed by indirecting through addresses into a store.
%
We borrow this technique on top of our own.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{\svar\mvar, \menv, \mstore, \mkont}$
    &
    $\tpl{\mval,\mstore,\mkont}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kcons{\kar{\mexpri1,\menv}}{\mkont}}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kcons{\kar{\mexpr,\menv}}{\mkont}}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kcons{\kfn{\mval}}{\mkont}}$
    \\
% function call
    $\tpl{\mval,\mstore,\kcons{\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}}}{\mkont}}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mval}, \mkont}$
  \end{tabular}
  \caption{The CESK machine with allocation tuning}
  \label{fig:base-semantics}
\end{figure}

\paragraph{Notation:} we make judicious use of the isomorphism between associations of $n$-tuples to spare the paper from excessive parentheses, even though our implementations are in a Lisp-descendent language.
